Four decades of water quality change in the upper San Francisco Estuary

### 2017-05-03 sftrends_manu.Rnw
final senn edits added

Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.  Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred, such as increased water clarity following a reduction of phytoplankton by biofiltration \cite{MacNally10}. The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}. We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
\end{inparaenum}

Invasion in the 1980s showed a clear displacement of the native \textit{Corbicula fluminea} with establishment of \textit{P. amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).  The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c).  A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).  An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).  These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (compare Figures \ref{fig:clmchl}e and \ref{fig:clmchl}e with \ref{fig:clmchl}d and \ref{fig:clmchl}g) such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.  Overall, these results suggest the increase in nutrient concentrations are from decreased uptake by phytoplankton in Suisun Bay.

@@

As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.  At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.  We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.  Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics. Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.  For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow-normalized results.  Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models).\cite{Beck17}  Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.

@@

<<clmchl, fig.height = 9.3, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.  Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).  A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \\ac{wrtds} results.  Reductions in \\ac{chla} concentration were coincident with an increase in SiO$_2$ and \\ac{din} concentrations (d, e), with the greatest increases in August (g, h). A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (i).  Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (j, observations from June).'>>=

>>>>

Invasion of the upper \ac{sfe} by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused severe changes in phytoplankton abundance and species composition.  Reduction in phytoplankton biomass has altered trophic networks in the upper \ac{sfe} and is considered an important mechanism in the decline of the protected delta smelt (\textit{Hypomesus transpacificus}) and other important fisheries \cite{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred, particularly increased water clarity from a reduction of particle transport and erodible sediment supply,\cite{Jassby08,Schoellhamer11,Cloern12b} although decreases in phytoplankton by clam biofiltration may have also increased clarity \cite{MacNally10}. The clams are halophilic such that drought years are correlated with an increase in biomass and further upstream invasion of the species \cite{Parchaso02,Cloern12b}. We hypothesized that results from \ac{wrtds} models would show \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, and
\item variation in the chlorophyll/clam relationship through indirect or direct controls of flow.
\end{inparaenum}
Although the relationship between phytoplankton and clams have been well described in \ac{sfe}\cite{Kimmerer14}, we use \ac{wrtds} to develop additional evidence that an increase in \ac{din} was facilitated in part by clam invasion.

Invasion in the 1980s showed a clear reduction of \textit{Corbicula fluminea} and increase of \textit{P. amurensis} (\cref{fig:clmchl}a), where biomass of the latter was negatively associated with flow from the Sacramento river (\Cref{fig:clmchl}b).  The increase in clam abundance was associated with a notable decrease in annually-averaged \ac{chla} from \ac{wrtds} results (\cref{fig:clmchl}c), as expected if \ac{wrtds} is adequately capturing flow variation and identifying the well-established phytoplankton decrease beginning in the 1980s.  A seasonal shift in the flow-normalized results was also observed such that \ac{chla} concentrations were generally highest in July/August prior to invasion, whereas a spring maximum in April was more common in recent years (\cref{fig:clmchl}f).  An increase in annually-averaged silicon dioxide (\cref{fig:clmchl}e) was coincident with the \ac{chla} decrease, with the largest increases occuring in August (\cref{fig:clmchl}g).  These relationships suggest that diatoms were the dominant genera early in the time series, particularly in late summer, whereas the spring peak observed in later years represents a shift to an earlier seasonal maxima.  This supports past research that showed a decrease in silica uptake by diatoms following invasion.\cite{Cloern96,Kimmerer05} Further, \ac{din} trends were similar to silicon-dioxide in both annual and seasonal changes (compare Figures \ref{fig:clmchl}e and \ref{fig:clmchl}e with \ref{fig:clmchl}d and \ref{fig:clmchl}g) such that an increase in both nutrients earlier in the time series corresponded with the decrease in \ac{chla}.  Overall, these results suggest that a nontrivial portion of the \ac{din} increase could be related to the decrease in a major `sink', i.e., decreased \ac{din} uptake by phytoplankton due to top down grazing pressure from \textit{P. amurensis}.

@@

As demonstrated by both case studies and the overall trends across all stations, water quality dynamics in the Delta are complex and driven by multiple factors that change through space and time.  At a minimum, \ac{wrtds} provides a description of change by focusing on high-level forcing factors that explicitly account for annual, seasonal, and flow effects on trend interpretations.  We have demonstrated the potential for imprecise or inaccurate conclusions of trend tests that focus solely on observed data and emphasize that flow-normalized trends have more power to quantify change.  Moreover, trends in nutrient loads from point soucres in the Delta have previously been described, e.g., Sacramento \ac{wwtp} increases\cite{Jassby08} and exports to Suisun Bay\cite{Novick14}. The results from \ac{wrtds} demonstrating these changes are not unexpected, and consequently, we are not detracting from the potential implications of such increases. The important conclusion is that the physical/hydrological and biogeochemical factors that influence nutrient cycling and ambient concentrations in the Bay-Delta, and changes to those factors, are substantial enough that they can be comparable in magnitude to anthropogenic load increases or comparable to the effects of management actions to decrease nutrient levels. Therefore, methods that adjust for the effects of these factors are critical when studying long-term records to assess the impacts or effectiveness of load increases or management actions, respectively.

Combined with additional data, \ac{wrtds} results can support hypotheses that lead to a more comprehensive understanding of ecosystem dynamics. Additional factors to consider include the effects of large-scale climatic patterns, more detailed hydrologic descriptions, and additional ecological components that affect trophic interactions.  For example, a more rigorous matching of flow time series with water quality observations at each station that considers varying source contributions over time could provide a more robust description of flow-normalized results.  Alternative methods for time series analysis could also be used to address a wider range of questions, particularly those with more generic structural forms that can explicitly include additional variables (e.g., generalized additive models).\cite{Beck17}  Overall, statistical interpretations of multiple factors can provide a basis for quantitative links between nutrient loads and adverse effects on ecosystem conditions, including the identification of thresholds for the protection and restoration of water quality.

@@

<<clmchl, fig.height = 9.3, fig.width = 8, out.width = '0.9\\textwidth', cache = T, fig.cap = 'Trends in clam abundance and \\ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.  Invasion by \\textit{Potamocorbula amurensis} clams in the late 1980s and reduction of \\textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).  A decrease in \\ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \\ac{wrtds} results.  Reductions in \\ac{chla} concentration were coincident with an increase in SiO$_2$ and \\ac{din} concentrations (d, e), with the greatest increases in August (g, h). A significant ($p < 0.001$) relationship between clam biomass and \\ac{chla} concentration is shown in subfigure (i).  Flow relationships with \\ac{chla} concentration shown by \\ac{wrtds} have also changed over time (j, observations from June).'>>=

### 2017-05-03 sftrends_manu.tex
final senn edits added

\caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.  Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and displacement of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).  A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \ac{wrtds} results.  Reductions in \ac{chla} concentration were coincident with an increase in SiO$_2$ and \ac{din} concentrations (d, e), with the greatest increases in August (g, h). A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (i).  Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (j, observations from June).}\label{fig:clmchl}
\end{figure}

>>>>

\caption{Trends in clam abundance and \ac{chla} concentration from 1976 to 2013 at station D7 in Suisun Bay.  Invasion by \textit{Potamocorbula amurensis} clams in the late 1980s and reduction of \textit{Corbicula fluminea} was shown by changes in clam density (a, annual means), with biomass linked to salinity (b).  A decrease in \ac{chla} concentration was also observed by changes in annual (c) and seasonal trends (f) based on \ac{wrtds} results.  Reductions in \ac{chla} concentration were coincident with an increase in SiO$_2$ and \ac{din} concentrations (d, e), with the greatest increases in August (g, h). A significant ($p < 0.001$) relationship between clam biomass and \ac{chla} concentration is shown in subfigure (i).  Flow relationships with \ac{chla} concentration shown by \ac{wrtds} have also changed over time (j, observations from June).}\label{fig:clmchl}
\end{figure}
